What We Eval in the Shadows

### 2021-08-11 paper/main.tex
Fix formatting of code blocks

\citet{ecoop12} gives a programming language-centric overview of the R language.
They characterized it as a lazy, vectorized, functional language with a rich
complement of dynamic features expressive enough to layer several object systems
on top of the core language. Most data types are sequences of primitive values.
For instance, \c{c("Ha","bye")} evaluates to a vector of two strings. Constants
such as \c{42} are vectors of length one. To enable equational reasoning, values
accessible through multiple aliases are copied when written to. Furthermore,
values can be tagged by attributes; these are key-value pairs. For instance, the
attribute \c{dim}-\c{c(2,2)} can be attached to the value at \c{x} by
\c{attr(x,"dim")$\leftarrow$c(2,2)}. Adding this attribute turns \c x into a
matrix. The \c{class} attribute gives a value a 'class' in the object-oriented
sense. So, \c{class(x)<-"human"} sets the class of \c{x} to \c{human}; classes
are used for method dispatch. Every linguistic construct is desugared to a
function call, even control flow statements, assignments and bracketing. All
functions can be shadowed and redefined, making R at the same time remarkably
flexible and exceedingly challenging to compile as vividly detailed
by~\citet{dls19}. R uses a relaxed call-by-need convention for passing arguments
to functions. Each argument is a thunk composed of an expression, its
environment, and a slot for the result; these are called \emph{promises}. To get
the value of an argument, the corresponding promise must be forced. Once forced,
the promise's result is cached for future use.

@@

The \eval function in R takes three parameters: an \c{expression} to evaluate
(\c{e}), an environment where to evaluate (\c{envir}), and an enclosure
(\c{encl}) that is used to look up objects not found in \c{envir}.\footnote{Next
to the \eval function R offers three other variants:
  \begin{inparaenum}[1.]
    \item \c{evalq} automatically quotes passed expression---a shorthand to
      \c{eval(quote(...))},
    \item \c{eval.parent(e, n)} specifies the evaluating environment in terms of
      number of call frames (\c{n}) to go back---a shorthand to \c{eval(e,
        parent.frame(n))}, and
    \item \c{local} evaluates \c{e} in a fresh environment---a shorthand to
      \c{evalq(e, new.env())}.
  \end{inparaenum} }

@@

\begin{figure}[htb]
\begin{minipage}{.49\textwidth}
\begin{lstlisting}[caption={Examples of calls producing expression \c{a+b}}]
  parse(text="a+b")
  quote(a+b)
  call("+", quote(a), quote(b))

@@

 The call to \c{substitute(x)} extracts the unevaluated expression from
the promise \c{x}. The function \c{match.call} returns an expression
representing the current call.

R is generous when it comes to what is allowed to be passed as environment to
\eval. Besides ordinary environments, it also accepts a list or a data
frame (using element names or column names for variable resolution) or an
integer \c{n} (in which case it would use the \c{n.} call frame). If no
environment is specified, then the default is the environment where the call to
\eval was made (as returned by \c{parent.frame}).
%
Environments nest, each has a parent. When creating a new environment with
\c{new.env}, the parent is the current environment. Environment chains can be
traversed with \c{parent.env}, until \c{emptyenv} is reached. The top-level
environment is \c{.GlobalEnv}, it has parents that represent the packages that
have been loaded. One can also directly read, modify or create new bindings,
given any environment:
%
\begin{lstlisting}
  # reading                                # writing
  envir$v                                  envir$v <- 2
  get("v", envir=envir)                    assign("v", 2, envir=envir)
\end{lstlisting}
%
 Environments are used as hash maps as they have reference semantics
and a built-in string lookup.

@@

\citet{wang} analyzed use of dynamic features in 18 Python programs to find if
they affect file change-proneness. Files with dynamic features are significantly
more likely to be the subject of changes than other files. Chen et al. looked
at the correlation between code changes and dynamic features, including \eval,
in 17 Python programs~\cite{chen}. They did not observe many uses of \eval.
\citet{oscar} performed an empirical study of the usage of dynamic features in
1,000 Smalltalk projects. While \eval itself is not present, Smalltalk has a
rich reflective interface. The authors found that reflective methods are used in less
than 2\% of methods. The most common reflective method is \c{perform:}; it send
a message that is specified by a string. These features are primarily used in the
core libraries.

@@

\citet{ecoop12} had a few paragraphs on the usage of \eval in R. They found the
that \eval is widely used in R code with 8500 call sites in 1238 CRAN packages
and 2.2 million dynamic calls. The 15 most frequent call sites account for 88\%
of those. The \c{match.arg} function is the highest used one with 54\% of all
calls. In the other call sites, they saw two use cases. The most common is the
evaluation of the source code of a promise retrieved by \c{substitute} in a new
environment, \eg as done in the \c{with} function. The other use case is the
invocation of a function whose name or arguments are determined dynamically. For
this purpose, R provides \c{do.call} and thus \eval is overkill. Our corpus is
larger, with \CranPackages packages, \packageTriggeredpkgs packages with at
least one \eval, for a total of \PkgHitEvalCallSites sites and
\packageAllcallsRnd dynamic calls, although they also looked at the Bioconductor
package repository in addition to CRAN. We also track much more runtime
information on the \eval calls during tracing. It enables us to have a more
nuanced and accurate picture, and identify more diverse use cases of \eval.

@@

\begin{compactenum}[(1)]
\item \emph{Download.} Packages are downloaded from CRAN, for Kaggle a web
  crawler retrieves code, and a command-line tool gets data. Installation is
  complicated by native dependencies which sometimes have to be resolved
  manually.
\item \emph{Extract.} The \genthat tool extracts all runnable code snippets and
  turns each of these into a self-standing program~\cite{issta18}, \c{knitr}
  extracts code from notebooks. Each extracted script is instrumented with calls
  to our dynamic analyzer in a way that avoid recording calls to \eval
  originating from our execution harness.
\item \emph{Trace.} Script run with a modified R interpreter to capture
  execution traces. Each script is run in its own process with the GNU R
  compiler turned off to avoid recording its execution. CRAN scripts run twice,
  once to capture \eval calls originating from package code and a second time to
  capture calls coming from base libraries.
\item \emph{Analyze.} Analysis outputs are merged, cleaned, and summarized.
  RMarkdown notebooks process the summarized data to generate all graphs (PDF
  from \c{ggplot2}) and numbers (exported as \LaTeX macros) appearing in the
  paper.
\end{compactenum}

@@

The tracer that performs dynamic analysis of R scripts is built on top of
\rdyntrace, an extended R 4.0.2 virtual machine that exposes callbacks for
various runtime events~\cite{oopsla19b}. The tracer registers callbacks to all
variants of \eval and a few additional functions to locate the origin of \eval
arguments (it taints the results of calls to \c{parse} and \c{match.call}). It
captures dynamic code loading, as well as variable definition and assignment,
allowing us to record side effects that happen in environments while evaluating
code in \eval. A challenge was that lack of source code references for
expression that are not within block surrounded by braces. This is unfortunately
not easily fixed. We extend our dynamic analysis tool to attach synthetic source
code references to all \eval call sites, but the approach fails in some edge
cases. For performance reasons, the tracer is an R package written in C++ (3.2K
LOC) and R (1.3K LOC). While in theory the implementation should be
straightforward, not so in practice. Lazy evaluation requires delaying the
processing of arguments until (and if) they are forced. Accounting for
side-effects performed in an \eval is complicated by the fact that R is
implemented in a mixture of R and C, and that the language implementation can
call \eval. Lastly, large code bases exercise many edge cases of the highly
underspecified R behavior.

@@

  \small
\begin{tabular}{@{}l@{\hspace{1.5cm}}l@{}}
\begin{minipage} {5cm}\small
  \begin{tabular}{r@{\,}r@{\,}l@{}r|r@{\,}r@{\,}l@{}r} \toprule
    \multicolumn{3}{c}{\bf \small\#calls} &\bf \small \#pck
&     \multicolumn{3}{c}{\bf \small\#calls} &\bf \small\#pck \\\midrule
\tt 1 &--& \tt 10      & \packageBina  & \tt 10K &--&\tt 100K  & \packageBine\\
\tt 11 &--& \tt 100    & \packageBinb  & \tt 100K &--&\tt 1M  & \packageBinf\\
\tt 101 &--& \tt 1K    & \packageBinc  & \tt 1M &--&\tt 10M   & \packageBing\\
\tt 1K &--& \tt 10K    & \packageBind  & \tt 10M &--& \tt 100M & \packageBinh\\\bottomrule
\end{tabular}
\caption{Call frequency}\label{freq}
\end{minipage}
&
\begin{minipage}{7cm}\small
\begin{tabular}{@{\,}r|rrrr}\toprule
  &\bf \eval & \bf \c{\bf evalq} & \c{\bf eval} & \c{\bf local}\\[-1.5mm]
           & & & \c{\bf .parent} &\\\midrule
\small Static sites &\packageStaticeval&\packageStaticevalq&\packageStaticevalparent&\packageStaticlocal \\
\small Exercised sites&\packageTriggeredeval&\packageTriggeredevalq&\packageTriggeredevalparent&\packageTriggeredlocal\\
\small Invocations&\packageEvalsRnd&\packageEvalqsRnd&\packageEparentsRnd&\packageLocalsRnd\\\bottomrule
\end{tabular}~\\[2mm]\caption{Variants}\label{tab:variantseval}
\end{minipage}\end{tabular}

@@

Our pipeline runs \CranRunnableScripts scripts extracted from \CranPackages
packages. Any run that does not call \eval is discarded, leaving us with
\packageNbruns runs. In these runs, \packageAllcalls \eval calls were recorded
originating from \PkgHitEvalCallSites sites. The runs exercised
\PkgHitEvalCallSitesAvgRatio of sites, a ratio similar to the package code
coverage metric which is \PkgCodeCoverage. The reasons some sites are not
exercised can be chalked down to incomplete tests and analysis failures
(\PkgFailedProgramsRatio of the runs crashed or timed out).
Fig.~\ref{fig:traced-eval-callsites} plots exercised site with respect to all
sites for each program; as can be seen coverage is unequal.
Fig.~\ref{freq} summarizes the frequency of calls, call counts on the left and
number of packages on the right. There are \packageFewcalls packages with fewer
than 100 calls, and \packageManycalls packages with more than 1,000 calls.
\packageMaxcallspack calls \eval \packageMaxcalls times and thus accounts for
over half of our observations.
Fig.~\ref{tab:variantseval} summarizes use of variants, the first row has sites
(\emph{static}), the second has sites encountered during analysis
(\emph{exercised}) and the last gives calls (\emph{invocations}). Most sites and
calls go to \eval itself, \c{eval.parent} is rare, and both \c{evalq} and
\c{local} are barely used at all.

@@

\Eval accepts any value as argument, but if the value is not an expression,
\eval returns it unchanged. Expressions account for \packageCodepercent of
arguments. More specifically, \packageSymbolpercent are \c{symbol}s (variables
such as \c{x}), \packageLanguagepercent are \c{language} objects (expressions
representing a function calls), and \packageExpressionpercent are \c{expression}
objects (lists of expressions). Further inspection reveals that \c{\_inherit}
accounts for \packageGgplotsymbolpercent of symbols and comes from one site in
\c{ggplot2}.\footnote{Each object has a field model inheritance, this comes from
\c{ggproto}, one of the many object-oriented systems for R.}

@@

To estimate how much executable code is injected through \eval, we measure the
size of the arguments in terms of number of nodes. For example,
\c{expression(x+1)} has a size of 3. The large number of symbols skew the median
to \packageMedianszeval but the average size is \packageAvgszeval nodes. The
largest \eval input we observed was an expression of \packageMaxszeval nodes, a
significant chunk of code. Fig.~\ref{fig:sizedistribution} shows the
distribution of sizes for arguments of fewer than 25 nodes. Sizes drop rapidly,
there are few observations larger than 15 nodes. The long tail is omitted for
legibility. As an alternative to counting nodes, we tried measuring string
lengths of expressions after calling a function to deparse values back to
strings, but these measurements were dominated by massive data objects which
could range in the MBs.

@@

\begin{table}[!b]\small
\begin{tabular}{crrrrrc}\toprule
  \bf $min(e)$& \bf \#sites & \bf \%sites & \bf \#packages & \bf \#operations & \bf \%envir & \bf example\\\midrule
\c X&\packageMinimizedcallsitesa &\packageMinimizedpropsitesa &\packageMinimizedpackagea &\packageMinimizedmedianoperationsaRnd &\packageMinimizedpercentparentframesa & \c{y+1}\\
\c{F(F(X))} & \packageMinimizedcallsitesb  & \packageMinimizedpropsitesb & \packageMinimizedpackageb  & \packageMinimizedmedianoperationsbRnd & \packageMinimizedpercentparentframesb & \c{gbov( mean(x), a-1)}\\
\c{V}&\packageMinimizedcallsitesc &\packageMinimizedpropsitesc &\packageMinimizedpackagec &\packageMinimizedmedianoperationscRnd &\packageMinimizedpercentparentframesc& \c{c(42,21,0)}\\
\c{F(X)}& \packageMinimizedcallsitesd & \packageMinimizedpropsitesd & \packageMinimizedpackaged & \packageMinimizedmedianoperationsdRnd & \packageMinimizedpercentparentframesd & \c{seq\_len(iters)} \\
\c{\$} & \packageMinimizedcallsitese & \packageMinimizedpropsitese & \packageMinimizedpackagee & \packageMinimizedmedianoperationseRnd & \packageMinimizedpercentparentframese & \c{DF\$B}\\
\c{model.frame}& \packageMinimizedcallsitesf & \packageMinimizedpropsitesf & \packageMinimizedpackagef & \packageMinimizedmedianoperationsfRnd & \packageMinimizedpercentparentframesf &  \c{model.frame(formula = Z $\sim$ U)}   \\
\c{F()}& \packageMinimizedcallsitesg & \packageMinimizedpropsitesg & \packageMinimizedpackageg & \packageMinimizedmedianoperationsgRnd & \packageMinimizedpercentparentframesg & \c{rgamma(3, 2, n = 10L)} \\
\c{FUN} & \packageMinimizedcallsitesh & \packageMinimizedpropsitesh & \packageMinimizedpackageh & \packageMinimizedmedianoperationshRnd & \packageMinimizedpercentparentframesh & \c{function(x, y) x + 3 * y} \\
\c{<-} & \packageMinimizedcallsitesi  & \packageMinimizedpropsitesi & \packageMinimizedpackagei & \packageMinimizedmedianoperationsiRnd & \packageMinimizedpercentparentframesi & \c{x[1, 2:3, 2:3] <- value}\\
\c{BLOCK} & \packageMinimizedcallsitesj & \packageMinimizedpropsitesj & \packageMinimizedpackagej & \packageMinimizedmedianoperationsjRnd & \packageMinimizedpercentparentframesj & \c{\{ write.csv(iris, tf) ; file.size(tf) \}} \\\bottomrule
\end{tabular}
\caption{Minimized expressions} \label{tab:minimizedexpressions}
\end{table}

The expressions passed to \eval vary widely. We categorize expressions using a
minimization function which abstract some of the incidental details of
expressions and return a ``shape'' that can be used to group similar \evals. The
function $min(e)$ which for a given expression $e$ returns a normal form. Normal
forms use \c V to stand in for values occurring in expression, \c X stands in
for variables, and \c F stands for functions. The function performs constant
folding of arithmetic and string expressions over base operators, for instance
$min(\c{1+1})$ simplifies to $\c{V}$, on the ground that addition of values
likely returns a value. Value simplification takes $min(\c{c(1,2,3+2)})$ to
$\c{V}$ as a complex vector containing values is a value. Variable absorption
has $min(\c{x+y})$ becomes $\c{X}$, this motivated by the fact that addition is
not an ``interesting'' operation, and that \c X stands in for any number of
variable lookups. Function absorption simplifies nested functions to keep an
abstraction of the nesting $min(\c{g(f(x),h(z))})=\c{F(F(X))}$. There are other
simplifications, the full list can be found in our artifact. It is worth
mentioning that these simplifications are heuristics, in R \c{1+1} is not
necessary \c 2. The addition operator can be redefined, but it typically isn't
or at least not in a way that would invalidate arithmetic.

@@

Contrary to JavaScript, in R the environment of \eval can be specifies by its
\c{env} argument. This gives users control over what is visible to the
computation started by \eval and the potential reach of its side-effects. We
distinguish the following kinds of environments:

\begin{itemize}[---]
\item \emph{Function:} environments for the local variables of some function
  currently active on the call stack. Obtained by calling \c{parent.frame()} or
  \c{sys.frame()}.
\item \emph{Synthetic:} environments built from data structures such as lists,
  data.frames, or constructed explicitly with \c{new.env}, \c{list2env}, or
  \c{as.environment} or the empty environment.
\item \emph{Global:} environments in which scripts or interactive commands are
  evaluated.
\item \emph{Package:} environments of libraries.
\end{itemize}

@@

\begin{itemize}[---]
\item {\it Reflection:} use of \c{match.call} to reflectively capture the
  expression that invoked this function, arguments are promises, and
  \c{substitute} is used to retrieve their source.
\item {\it String:} create from strings by invoking \c{parse},
  \c{str2expression}, or \c{str2lang}.
\item {\it Constructed:} invoked \c{quote} or \c{expression}, or by built with
  \c{call} or \c{as.call}.
\item {\it Environment: } created with \c{as.name} or \c{as.symbol}, typically
  to read a non-local environment.
\item {\it External: }  call to a C function. % Put an example
\end{itemize}

Table~\ref{tab:provenance} summarizes provenance. Some sites are invoked with
different provenances, they are counted multiple times. Strings correlate with
dynamic code loading as done by \c{source} and \c{sys.source}. Few calls
(\packageNbParseFromFileSites in total) consume the result of calling \c{parse}
on a file. Most build strings programmatically. We identified one function
\c{invokeRestartInteractively} that prompts the user for input and \evals it.

\paragraph{Discussion}
The main use case for \eval involves using \c{match.call}, \c{substitute}, or
\c{formals} to access and transform arguments of a function. Constructing
expressions from strings is also a common use case. The constructed category
shows that most \evals comes from existing code that was modified by the
programmer before invoking \eval. Both constructed and reflection categories
roughly correspond to meta-programming. Some use cases could likely be replaced
by macros if the R designers could be convinced to overcome their distaste for
those. Another use of \eval is to directly build symbols to evaluate them in
another environment.

>>>>

\citet{ecoop12} gives a programming language-centric overview of the R language.
They characterized it as a lazy, vectorized, functional language with a rich
complement of dynamic features expressive enough to layer several object systems
on top of the core language. Most data types are sequences of primitive values.
For instance, \code{c("Ha","bye")} evaluates to a vector of two strings. Constants
such as \code{42} are vectors of length one. To enable equational reasoning, values
accessible through multiple aliases are copied when written to. Furthermore,
values can be tagged by attributes; these are key-value pairs. For instance, the
attribute \code{dim}-\code{c(2,2)} can be attached to the value at \code{x} by
\code{attr(x,"dim") <- c(2,2)}. Adding this attribute turns \code{x} into a
matrix. The \code{class} attribute gives a value a 'class' in the object-oriented
sense. So, \code{class(x)<-"human"} sets the class of \code{x} to \code{human}; classes
are used for method dispatch. Every linguistic construct is desugared to a
function call, even control flow statements, assignments and bracketing. All
functions can be shadowed and redefined, making R at the same time remarkably
flexible and exceedingly challenging to compile as vividly detailed
by~\citet{dls19}. R uses a relaxed call-by-need convention for passing arguments
to functions. Each argument is a thunk composed of an expression, its
environment, and a slot for the result; these are called \emph{promises}. To get
the value of an argument, the corresponding promise must be forced. Once forced,
the promise's result is cached for future use.

@@

The \eval function in R takes three parameters: an \code{expression} to evaluate
(\code{e}), an environment where to evaluate (\code{envir}), and an enclosure
(\code{encl}) that is used to look up objects not found in \code{envir}.\footnote{Next
to the \eval function R offers three other variants:
  \begin{inparaenum}[1.]
    \item \code{evalq} automatically quotes passed expression---a shorthand to
      \code{eval(quote(...))},
    \item \code{eval.parent(e, n)} specifies the evaluating environment in terms of
      number of call frames (\code{n}) to go back---a shorthand to \code{eval(e,
        parent.frame(n))}, and
    \item \code{local} evaluates \code{e} in a fresh environment---a shorthand to
      \code{evalq(e, new.env())}.
  \end{inparaenum} }

@@

\begin{figure}[htb]
\begin{minipage}{.49\textwidth}
\begin{lstlisting}[caption={Examples of calls producing expression \code{a+b}}]
  parse(text="a+b")
  quote(a+b)
  call("+", quote(a), quote(b))

@@

 The call to \code{substitute(x)} extracts the unevaluated expression from
the promise \code{x}. The function \code{match.call} returns an expression
representing the current call.

R is generous when it comes to what is allowed to be passed as environment to
\eval. Besides ordinary environments, it also accepts a list or a data
frame (using element names or column names for variable resolution) or an
integer \code{n} (in which case it would use the \code{n.} call frame). If no
environment is specified, then the default is the environment where the call to
\eval was made (as returned by \code{parent.frame}).
%
Environments nest, each has a parent. When creating a new environment with
\code{new.env}, the parent is the current environment. Environment chains can be
traversed with \code{parent.env}, until \code{emptyenv} is reached. The top-level
environment is \code{.GlobalEnv}, it has parents that represent the packages that
have been loaded. One can also directly read, modify or create new bindings,
given any environment:
%
\begin{lstlisting}
  # reading                                # writing
  envir$v                                  envir$v <- 2
  get("v", envir=envir)                    assign("v", 2, envir=envir)
\end{lstlisting}
%
 Environments are used as hash maps as they have reference semantics
and a built-in string lookup.

@@

\citet{wang} analyzed use of dynamic features in 18 Python programs to find if
they affect file change-proneness. Files with dynamic features are significantly
more likely to be the subject of changes than other files. Chen et al. looked
at the correlation between code changes and dynamic features, including \eval,
in 17 Python programs~\cite{chen}. They did not observe many uses of \eval.
\citet{oscar} performed an empirical study of the usage of dynamic features in
1,000 Smalltalk projects. While \eval itself is not present, Smalltalk has a
rich reflective interface. The authors found that reflective methods are used in less
than 2\% of methods. The most common reflective method is \code{perform:}; it send
a message that is specified by a string. These features are primarily used in the
core libraries.

@@

\citet{ecoop12} had a few paragraphs on the usage of \eval in R. They found the
that \eval is widely used in R code with 8500 call sites in 1238 CRAN packages
and 2.2 million dynamic calls. The 15 most frequent call sites account for 88\%
of those. The \code{match.arg} function is the highest used one with 54\% of all
calls. In the other call sites, they saw two use cases. The most common is the
evaluation of the source code of a promise retrieved by \code{substitute} in a new
environment, \eg as done in the \code{with} function. The other use case is the
invocation of a function whose name or arguments are determined dynamically. For
this purpose, R provides \code{do.call} and thus \eval is overkill. Our corpus is
larger, with \CranPackages packages, \packageTriggeredpkgs packages with at
least one \eval, for a total of \PkgHitEvalCallSites sites and
\packageAllcallsRnd dynamic calls, although they also looked at the Bioconductor
package repository in addition to CRAN. We also track much more runtime
information on the \eval calls during tracing. It enables us to have a more
nuanced and accurate picture, and identify more diverse use cases of \eval.

@@

\begin{compactenum}[(1)]
\item \emph{Download.} Packages are downloaded from CRAN, for Kaggle a web
  crawler retrieves code, and a command-line tool gets data. Installation is
  complicated by native dependencies which sometimes have to be resolved
  manually.
\item \emph{Extract.} The \genthat tool extracts all runnable code snippets and
  turns each of these into a self-standing program~\cite{issta18}, \code{knitr}
  extracts code from notebooks. Each extracted script is instrumented with calls
  to our dynamic analyzer in a way that avoid recording calls to \eval
  originating from our execution harness.
\item \emph{Trace.} Script run with a modified R interpreter to capture
  execution traces. Each script is run in its own process with the GNU R
  compiler turned off to avoid recording its execution. CRAN scripts run twice,
  once to capture \eval calls originating from package code and a second time to
  capture calls coming from base libraries.
\item \emph{Analyze.} Analysis outputs are merged, cleaned, and summarized.
  RMarkdown notebooks process the summarized data to generate all graphs (PDF
  from \code{ggplot2}) and numbers (exported as \LaTeX macros) appearing in the
  paper.
\end{compactenum}

@@

The tracer that performs dynamic analysis of R scripts is built on top of
\rdyntrace, an extended R 4.0.2 virtual machine that exposes callbacks for
various runtime events~\cite{oopsla19b}. The tracer registers callbacks to all
variants of \eval and a few additional functions to locate the origin of \eval
arguments (it taints the results of calls to \code{parse} and \code{match.call}). It
captures dynamic code loading, as well as variable definition and assignment,
allowing us to record side effects that happen in environments while evaluating
code in \eval. A challenge was that lack of source code references for
expression that are not within block surrounded by braces. This is unfortunately
not easily fixed. We extend our dynamic analysis tool to attach synthetic source
code references to all \eval call sites, but the approach fails in some edge
cases. For performance reasons, the tracer is an R package written in C++ (3.2K
LOC) and R (1.3K LOC). While in theory the implementation should be
straightforward, not so in practice. Lazy evaluation requires delaying the
processing of arguments until (and if) they are forced. Accounting for
side-effects performed in an \eval is complicated by the fact that R is
implemented in a mixture of R and C, and that the language implementation can
call \eval. Lastly, large code bases exercise many edge cases of the highly
underspecified R behavior.

@@

  \small
\begin{tabular}{@{}l@{\hspace{1.5cm}}l@{}}
\begin{minipage} {5cm}\small
  \begin{tabular}{r@{\,}r@{\,}l@{}r|r@{\,}r@{\,}l@{}r} \toprule
    \multicolumn{3}{c}{\bf \small\#calls} &\bf \small \#pck
&     \multicolumn{3}{c}{\bf \small\#calls} &\bf \small\#pck \\\midrule
\tt 1 &--& \tt 10      & \packageBina  & \tt 10K &--&\tt 100K  & \packageBine\\
\tt 11 &--& \tt 100    & \packageBinb  & \tt 100K &--&\tt 1M  & \packageBinf\\
\tt 101 &--& \tt 1K    & \packageBinc  & \tt 1M &--&\tt 10M   & \packageBing\\
\tt 1K &--& \tt 10K    & \packageBind  & \tt 10M &--& \tt 100M & \packageBinh\\\bottomrule
\end{tabular}
\caption{Call frequency}\label{freq}
\end{minipage}
&
\begin{minipage}{7cm}\small
\begin{tabular}{@{\,}r|rrrr}\toprule
  &\eval & \code{evalq} & \code{eval} & \code{local}\\[-1.5mm]
           & & & \code{.parent} &\\\midrule
\small Static sites &\packageStaticeval&\packageStaticevalq&\packageStaticevalparent&\packageStaticlocal \\
\small Exercised sites&\packageTriggeredeval&\packageTriggeredevalq&\packageTriggeredevalparent&\packageTriggeredlocal\\
\small Invocations&\packageEvalsRnd&\packageEvalqsRnd&\packageEparentsRnd&\packageLocalsRnd\\\bottomrule
\end{tabular}~\\[2mm]\caption{Variants}\label{tab:variantseval}
\end{minipage}\end{tabular}

@@

Our pipeline runs \CranRunnableScripts scripts extracted from \CranPackages
packages. Any run that does not call \eval is discarded, leaving us with
\packageNbruns runs. In these runs, \packageAllcalls \eval calls were recorded
originating from \PkgHitEvalCallSites sites. The runs exercised
\PkgHitEvalCallSitesAvgRatio of sites, a ratio similar to the package code
coverage metric which is \PkgCodeCoverage. The reasons some sites are not
exercised can be chalked down to incomplete tests and analysis failures
(\PkgFailedProgramsRatio of the runs crashed or timed out).
Fig.~\ref{fig:traced-eval-callsites} plots exercised site with respect to all
sites for each program; as can be seen coverage is unequal.
Fig.~\ref{freq} summarizes the frequency of calls, call counts on the left and
number of packages on the right. There are \packageFewcalls packages with fewer
than 100 calls, and \packageManycalls packages with more than 1,000 calls.
\packageMaxcallspack calls \eval \packageMaxcalls times and thus accounts for
over half of our observations.
Fig.~\ref{tab:variantseval} summarizes use of variants, the first row has sites
(\emph{static}), the second has sites encountered during analysis
(\emph{exercised}) and the last gives calls (\emph{invocations}). Most sites and
calls go to \eval itself, \code{eval.parent} is rare, and both \code{evalq} and
\code{local} are barely used at all.

@@

\Eval accepts any value as argument, but if the value is not an expression,
\eval returns it unchanged. Expressions account for \packageCodepercent of
arguments. More specifically, \packageSymbolpercent are \code{symbol}s (variables
such as \code{x}), \packageLanguagepercent are \code{language} objects (expressions
representing a function calls), and \packageExpressionpercent are \code{expression}
objects (lists of expressions). Further inspection reveals that \code{\_inherit}
accounts for \packageGgplotsymbolpercent of symbols and comes from one site in
\code{ggplot2}.\footnote{Each object has a field model inheritance, this comes from
\code{ggproto}, one of the many object-oriented systems for R.}

@@

To estimate how much executable code is injected through \eval, we measure the
size of the arguments in terms of number of nodes. For example,
\code{expression(x+1)} has a size of 3. The large number of symbols skew the median
to \packageMedianszeval but the average size is \packageAvgszeval nodes. The
largest \eval input we observed was an expression of \packageMaxszeval nodes, a
significant chunk of code. Fig.~\ref{fig:sizedistribution} shows the
distribution of sizes for arguments of fewer than 25 nodes. Sizes drop rapidly,
there are few observations larger than 15 nodes. The long tail is omitted for
legibility. As an alternative to counting nodes, we tried measuring string
lengths of expressions after calling a function to deparse values back to
strings, but these measurements were dominated by massive data objects which
could range in the MBs.

@@

\begin{table}[!b]\small
\begin{tabular}{crrrrrc}\toprule
  \bf $min(e)$& \bf \#sites & \bf \%sites & \bf \#packages & \bf \#operations & \bf \%envir & \bf example\\\midrule
\code{X}&\packageMinimizedcallsitesa &\packageMinimizedpropsitesa &\packageMinimizedpackagea &\packageMinimizedmedianoperationsaRnd &\packageMinimizedpercentparentframesa & \code{y+1}\\
\code{F(F(X))} & \packageMinimizedcallsitesb  & \packageMinimizedpropsitesb & \packageMinimizedpackageb  & \packageMinimizedmedianoperationsbRnd & \packageMinimizedpercentparentframesb & \code{gbov( mean(x), a-1)}\\
\code{V}&\packageMinimizedcallsitesc &\packageMinimizedpropsitesc &\packageMinimizedpackagec &\packageMinimizedmedianoperationscRnd &\packageMinimizedpercentparentframesc& \code{c(42,21,0)}\\
\code{F(X)}& \packageMinimizedcallsitesd & \packageMinimizedpropsitesd & \packageMinimizedpackaged & \packageMinimizedmedianoperationsdRnd & \packageMinimizedpercentparentframesd & \code{seq\_len(iters)} \\
\code{\$} & \packageMinimizedcallsitese & \packageMinimizedpropsitese & \packageMinimizedpackagee & \packageMinimizedmedianoperationseRnd & \packageMinimizedpercentparentframese & \code{DF\$B}\\
\code{model.frame}& \packageMinimizedcallsitesf & \packageMinimizedpropsitesf & \packageMinimizedpackagef & \packageMinimizedmedianoperationsfRnd & \packageMinimizedpercentparentframesf &  \code{model.frame(formula = Z ~ U)}   \\
\code{F()}& \packageMinimizedcallsitesg & \packageMinimizedpropsitesg & \packageMinimizedpackageg & \packageMinimizedmedianoperationsgRnd & \packageMinimizedpercentparentframesg & \code{rgamma(3, 2, n = 10L)} \\
\code{FUN} & \packageMinimizedcallsitesh & \packageMinimizedpropsitesh & \packageMinimizedpackageh & \packageMinimizedmedianoperationshRnd & \packageMinimizedpercentparentframesh & \code{function(x, y) x + 3 * y} \\
\code{<-} & \packageMinimizedcallsitesi  & \packageMinimizedpropsitesi & \packageMinimizedpackagei & \packageMinimizedmedianoperationsiRnd & \packageMinimizedpercentparentframesi & \code{x[1, 2:3, 2:3] <- value}\\
\code{BLOCK} & \packageMinimizedcallsitesj & \packageMinimizedpropsitesj & \packageMinimizedpackagej & \packageMinimizedmedianoperationsjRnd & \packageMinimizedpercentparentframesj & \code{\{write.csv(iris,tf);file.size(tf)\}} \\\bottomrule
\end{tabular}
\caption{Minimized expressions} \label{tab:minimizedexpressions}
\end{table}

The expressions passed to \eval vary widely. We categorize expressions using a
minimization function which abstract some of the incidental details of
expressions and return a ``shape'' that can be used to group similar \evals. The
function $min(e)$ which for a given expression $e$ returns a normal form. Normal
forms use \code{V} to stand in for values occurring in expression, \code{X} stands in
for variables, and \code{F} stands for functions. The function performs constant
folding of arithmetic and string expressions over base operators, for instance
$min(\c{1+1})$ simplifies to $\c{V}$, on the ground that addition of values
likely returns a value. Value simplification takes $min(\c{c(1,2,3+2)})$ to
$\c{V}$ as a complex vector containing values is a value. Variable absorption
has $min(\c{x+y})$ becomes $\c{X}$, this motivated by the fact that addition is
not an ``interesting'' operation, and that \code{X} stands in for any number of
variable lookups. Function absorption simplifies nested functions to keep an
abstraction of the nesting $min(\c{g(f(x),h(z))})=\c{F(F(X))}$. There are other
simplifications, the full list can be found in our artifact. It is worth
mentioning that these simplifications are heuristics, in R \code{1+1} is not
necessary \code{2}. The addition operator can be redefined, but it typically isn't
or at least not in a way that would invalidate arithmetic.

@@

Contrary to JavaScript, in R the environment of \eval can be specifies by its
\code{env} argument. This gives users control over what is visible to the
computation started by \eval and the potential reach of its side-effects. We
distinguish the following kinds of environments:

\begin{itemize}[---]
\item \emph{Function:} environments for the local variables of some function
  currently active on the call stack. Obtained by calling \code{parent.frame()} or
  \code{sys.frame()}.
\item \emph{Synthetic:} environments built from data structures such as lists,
  data.frames, or constructed explicitly with \code{new.env}, \code{list2env}, or
  \code{as.environment} or the empty environment.
\item \emph{Global:} environments in which scripts or interactive commands are
  evaluated.
\item \emph{Package:} environments of libraries.
\end{itemize}

@@

\begin{itemize}[---]
\item {\it Reflection:} use of \code{match.call} to reflectively capture the
  expression that invoked this function, arguments are promises, and
  \code{substitute} is used to retrieve their source.
\item {\it String:} create from strings by invoking \code{parse},
  \code{str2expression}, or \code{str2lang}.
\item {\it Constructed:} invoked \code{quote} or \code{expression}, or by built with
  \code{call} or \code{as.call}.
\item {\it Environment: } created with \code{as.name} or \code{as.symbol}, typically
  to read a non-local environment.
\item {\it External: }  call to a C function. % Put an example
\end{itemize}

Table~\ref{tab:provenance} summarizes provenance. Some sites are invoked with
different provenances, they are counted multiple times. Strings correlate with
dynamic code loading as done by \code{source} and \code{sys.source}. Few calls
(\packageNbParseFromFileSites in total) consume the result of calling \code{parse}
on a file. Most build strings programmatically. We identified one function
\code{invokeRestartInteractively} that prompts the user for input and \evals it.

\paragraph{Discussion}
The main use case for \eval involves using \code{match.call}, \code{substitute}, or
\code{formals} to access and transform arguments of a function. Constructing
expressions from strings is also a common use case. The constructed category
shows that most \evals comes from existing code that was modified by the
programmer before invoking \eval. Both constructed and reflection categories
roughly correspond to meta-programming. Some use cases could likely be replaced
by macros if the R designers could be convinced to overcome their distaste for
those. Another use of \eval is to directly build symbols to evaluate them in
another environment.
